### 2024-04-19 docs/SOSP 2024/exploring.tex
Shrink tables and algorithm listing
Shrink some figures
Add representation values to table
pass

\subsection{Bounding the search space}
We bound the space using a few meta-parameters that together define 
the primitives used to construct a schedule. We do so while making sure 
that the space (at-least) covers the known strategies published in prior work.
Specifically, we use 4 numeric parameters and 4 boolean parameters listed in Table \ref{tab:schedparams}. 
%Note that even this template has a large number ($1728$) of schedules, and requires further pruning as we discuss later.
The first three numeric parameters assign a configurable number of rows to each 
thread block, to each thread, and determine how trees are distributed across a
specified number of threads. These parameters together define the loop schedule 
primitives to use (including the arguments to pass) from Table~\ref{Tab:LoopModifiers} 
and the parallelization strategy. The next four parameters determine the caching strategy, 
unroll factor{\footnote{We choose to always unroll trees completely, while its possible 
to partially unroll loops we did not observe any performance benefits from doing so.}} 
and how many trees to traverse simultaneously \TODO{Should we say, in the kernel/within
a single thread?}.  (the remaining primitives from Table~\ref{Tab:Optimizations}).  
The final parameter determines what reduction options to use (Table~\ref{Tab:ReductionOpts}).
%Table \ref{tab:schedparams} lists the parameter values we tried. 
% \begin{itemize}
%   \item \textbf{Number of rows per thread block (Integer):} The number of rows that are processed by each thread block.
%   \item \textbf{Number of rows per thread (Integer):} The number of rows processed by each thread.
%   \item \textbf{Number of tree threads (Integer):} The number of threads across which the trees are distributed.
%   \item \textbf{Cache rows (Boolean):} Whether the input rows are cached in shared memory.
%   \item \textbf{Cache trees (Boolean):} Whether the trees are cached in shared memory.
%   \item \textbf{Unroll walks (Boolean):} Whether the tree walks are unrolled.
%   \item \textbf{Tree walk interleave factor:} The number of tree walks that are interleaved.
%   \item \textbf{Shared memory reduction:} Whether the reduction across tree threads is done in shared memory.
% \end{itemize}

\begin{table}[htb]
  \centering
  \resizebox{\linewidth}{!}{
  \begin{tabularx}{\linewidth}{c | c }
  \toprule
  \textbf{Parameter} & \textbf{Values} \\
  \midrule
  \textbf{Rows per thread block} & $\{8, 32, 64\}$ \\
  \textbf{Rows per thread} & $\{1, 2, 4\}$ \\
  \textbf{Number of tree threads} & $\{2, 10, 20, 50\}$ \\
  \textbf{Cache rows} & $\{True, False\}$ \\
  \textbf{Cache trees} & $\{True, False\}$ \\
  \textbf{Unroll walks} & $\{True, False\}$ \\
  \textbf{Tree walk interleave factor} & $\{1, 2, 4\}$ \\
  \textbf{Shared memory reduction} & $\{True, False\}$ \\
  \bottomrule
  \end{tabularx}
  }
  \vskip 5pt
  \caption{\label{tab:schedparams} List of parameter values we explored for the template GPU schedule.}
\end{table}

@@

\begin{figure}[htb]
  \centering
  \includegraphics[width=\linewidth]{figures/normalized_kernel_histogram_lt5.png}
  \caption{Distribution of normalized execution times for all benchmark models
  with the template schedule using parameter values as shown in Table \ref{tab:schedparams}}. 
  \label{Fig:ExecTimeDistribution}
\end{figure}

We evaluated all the schedules in this space on several real world models and observed that 
there is huge variation in performance with different parameter values. 
Figure \ref{Fig:ExecTimeDistribution} shows the distribution of normalized execution times
for some real-world models with different parameter values for the template schedule (inference
times normalized w.r.t fastest time for that model).
As can be seen very few schedules perform close to the best while a vast majority of
schedules perform poorly.

Exhaustive exploration over this bounded space ($5184$ schedules) is still very expensive, taking
anywhere between thirty minutes to a few hours to explore the entire space for a given model. 
%Exploring the space for every model is infeasible.
%\TODO{comment on online vs offline exploration? Set a goal for the heuristic?}

@@

\subsubsection*{Insufficient parallelism}
 Some configurations do not expose sufficient parallelism, we prune these out.
  For example when the batch size is small, it does not make sense to have a many rows
  per thread block or to partition the trees across a few threads. 
  We therefore limit the combinations used based on batch size (see lines 3---8).
\subsubsection*{Utilizing shared memory}
  Shared memory is a critical resource on GPUs and it helps to only bring objects that would be reused into it. 
  We observe that tree nodes have limited reuse and the one time cost of loading trees is not sufficiently amortized when the whole tree is not accessed during inference. 
  On the other hand caching rows 
  %when possible (i.e., when the number of features is small enough to fit in shared memory)
  almost always improves performance. Further when the inputs have many features, it may not be possible to cache all rows.
  In this scenario it helps to retain a few rows in memory, and pick schedules similar to the small batch size case (lines 15,3---5).

@@

\begin{algorithm}
  \caption{Heuristic to find a good schedule}
  \label{alg:heuristic}
  \begin{algorithmic}[1]
  \Procedure{TBConfigs}{$N_{batch}$, $N_f$}
    \State $T_{batch} \gets 2048,\; T_f \gets 128$
    \If {$N_{batch} \leq T_{batch}$ \textbf{or} $N_{f} > T_{f}$}
      \State $rowsPerBlock \gets \{8, 32\}$
      \State $treeThreads \gets \{20, 50\}$
    \Else
      \State $rowsPerBlock \gets \{32, 64\}$
      \State $treeThreads \gets \{2, 10\}$
    \EndIf
    \State \Return $rowsPerBlock, treeThreads$
  \EndProcedure
  \\
  \State $bestSchedules \gets shMemSchedules \gets \emptyset$
  \State $rowsPerTB, treeThds \gets TBConfigs(N_{batch}, N_f)$
  \State $cacheRows \gets \text{True},\; cacheTrees \gets \text{False}$
  \State $interleave \gets \{1, 2, 4\}$
  % \State $reps \gets \{array, sparse, reorg\}$
  \State $schedules \gets (rowsPerTB, treeThds, cacheRows,$ \\
                          \hspace{2cm}$cacheTrees, interleave)$
                          %\hspace{2cm}$$
  \For{$(sched, rep) \in schedules \times \{array, sparse, reorg\}$}
    \State $time \gets EvaluateSchedule(sched, rep)$
    \State $bestSchedules.insert(time, sched, rep)$
  \EndFor
  \State 
  \For{$sched, rep \in Top3(bestSchedules)$}
    \State $EnableSharedReduction(sched)$
    \State $time \gets EvaluateSchedule(sched, rep)$
    \State $shMemSchedules.insert(time, sched, rep)$
  \EndFor
  \State \Return $min(shMemSchedules \cup bestSchedules)$
  \end{algorithmic} 
\end{algorithm}

>>>>

\subsection{Bounding the search space}
We bound the space using a few meta-parameters that together define 
the primitives used to construct a schedule. We do so while making sure 
that the space (at-least) covers the known strategies published in prior work.
Specifically, we use 4 numeric parameters, 4 boolean parameters and one ternary parameter as listed in Table \ref{tab:schedparams}.
We picked the specific values listed through experimentation.
%Note that even this template has a large number ($1728$) of schedules, and requires further pruning as we discuss later.
The first three numeric parameters assign a configurable number of rows to each 
thread block, to each thread, and determine how trees are distributed across a
specified number of threads. These parameters together define the loop schedule 
primitives to use (including the arguments to pass) from Table~\ref{Tab:LoopModifiers} 
and the parallelization strategy. The next four parameters determine the caching strategy, 
unroll factor{\footnote{We choose to always unroll trees completely, while its possible 
to partially unroll loops we did not observe any performance benefits from doing so.}} 
and how many trees to traverse simultaneously \TODO{Should we say, in the kernel/within
a single thread?}.  (the remaining primitives from Table~\ref{Tab:Optimizations}).  
The last two parameters determine reduction option (Table~\ref{Tab:ReductionOpts}) and the layout{\footnote{This does not have a schedule primitive. We always explore all three layout options.}} to use.
%Table \ref{tab:schedparams} lists the parameter values we tried. 
% \begin{itemize}
%   \item \textbf{Number of rows per thread block (Integer):} The number of rows that are processed by each thread block.
%   \item \textbf{Number of rows per thread (Integer):} The number of rows processed by each thread.
%   \item \textbf{Number of tree threads (Integer):} The number of threads across which the trees are distributed.
%   \item \textbf{Cache rows (Boolean):} Whether the input rows are cached in shared memory.
%   \item \textbf{Cache trees (Boolean):} Whether the trees are cached in shared memory.
%   \item \textbf{Unroll walks (Boolean):} Whether the tree walks are unrolled.
%   \item \textbf{Tree walk interleave factor:} The number of tree walks that are interleaved.
%   \item \textbf{Shared memory reduction:} Whether the reduction across tree threads is done in shared memory.
% \end{itemize}

\begin{table}[htb]
  \footnotesize
  \centering
  % \resizebox{\linewidth}{!}{
  \begin{tabularx}{\linewidth}{c | c }
    \toprule
    \textbf{Parameter} & \textbf{Values} \\
    \midrule
    \textbf{Rows per thread block} & $\{8, 32, 64\}$ \\
    \textbf{Rows per thread} & $\{1, 2, 4\}$ \\
    \textbf{Number of tree threads} & $\{2, 10, 20, 50\}$ \\
    \textbf{Cache rows} & $\{True, False\}$ \\
    \textbf{Cache trees} & $\{True, False\}$ \\
    \textbf{Unroll walks} & $\{True, False\}$ \\
    \textbf{Tree walk interleave factor} & $\{1, 2, 4\}$ \\
    \textbf{Shared memory reduction} & $\{True, False\}$ \\
    \textbf{Representation} & $\{array, sparse, reorg\}$ \\
    \bottomrule
  \end{tabularx}
  % }
  \vskip 5pt
  \caption{\label{tab:schedparams} List of parameter values we explored for the template GPU schedule.}
\end{table}

@@

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.7\linewidth]{figures/normalized_kernel_histogram_lt5.png}
  \caption{Distribution of normalized execution times for all benchmark models
  with the template schedule using parameter values as shown in Table \ref{tab:schedparams}}. 
  \label{Fig:ExecTimeDistribution}
\end{figure}

We evaluated all the schedules in this space on several real world models and observed that 
there is huge variation in performance with different parameter values. 
Figure \ref{Fig:ExecTimeDistribution} shows the distribution of normalized execution times
for some real-world models with different parameter values for the template schedule (inference
times normalized w.r.t fastest time for that model).
As can be seen very few schedules perform close to the best while a vast majority perform poorly.

Exhaustive exploration over this bounded space ($5184$ schedules) is still very expensive, taking
 between $30-600$ minutes to explore the entire space for a given model. 
%Exploring the space for every model is infeasible.
%\TODO{comment on online vs offline exploration? Set a goal for the heuristic?}

@@

\subsubsection*{Insufficient parallelism}
 Some configurations do not expose sufficient parallelism, we prune these out.
  For example when the batch size is small, it does not make sense to have many rows
  per thread block or to partition the trees across a few threads. 
  We therefore limit the combinations used based on batch size (see lines 3---8).
\subsubsection*{Utilizing shared memory}
  Shared memory is a critical resource on GPUs and it helps to only bring objects that would be reused into it. 
  We observe that tree nodes have limited reuse and the one time cost of loading trees is not sufficiently amortized when the whole tree is not accessed during inference. 
  On the other hand caching rows 
  %when possible (i.e., when the number of features is small enough to fit in shared memory)
  almost always improves performance. Further when the inputs have many features, it may not be possible to cache all rows.
  In this scenario it helps to retain a few rows in memory, and pick schedules similar to the small batch size case (lines 15,3---5).

@@

\begin{algorithm}
  \caption{Heuristic to find a good schedule}
  \label{alg:heuristic}
  \small{
  \begin{algorithmic}[1]
  \Procedure{TBConfigs}{$N_{batch}$, $N_f$}
    \State $T_{batch} \gets 2048,\; T_f \gets 128$
    \If {$N_{batch} \leq T_{batch}$ \textbf{or} $N_{f} > T_{f}$}
      \State $rowsPerBlock \gets \{8, 32\}$
      \State $treeThreads \gets \{20, 50\}$
    \Else
      \State $rowsPerBlock \gets \{32, 64\}$
      \State $treeThreads \gets \{2, 10\}$
    \EndIf
    \State \Return $rowsPerBlock, treeThreads$
  \EndProcedure
  \\
  \State $bestSchedules \gets shMemSchedules \gets \emptyset$
  \State $rowsPerTB, treeThds \gets TBConfigs(N_{batch}, N_f)$
  \State $cacheRows \gets \text{True},\; cacheTrees \gets \text{False}$
  \State $interleave \gets \{1, 2, 4\}$
  % \State $reps \gets \{array, sparse, reorg\}$
  \State $schedules \gets (rowsPerTB, treeThds, cacheRows,$ \\
                          \hspace{2cm}$cacheTrees, interleave)$
                          %\hspace{2cm}$$
  \For{$(sched, rep) \in schedules \times \{array, sparse, reorg\}$}
    \State $time \gets EvaluateSchedule(sched, rep)$
    \State $bestSchedules.insert(time, sched, rep)$
  \EndFor
  \State 
  \For{$sched, rep \in Top3(bestSchedules)$}
    \State $EnableSharedReduction(sched)$
    \State $time \gets EvaluateSchedule(sched, rep)$
    \State $shMemSchedules.insert(time, sched, rep)$
  \EndFor
  \State \Return $min(shMemSchedules \cup bestSchedules)$
  \end{algorithmic} 
  }
\end{algorithm}
